Abstract Binding Trees

### 2015-10-22 paper.tex
implement variable substitution (close #2)
Fix the category of symbol contexts

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath, amsthm, stmaryrd, scalerel}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{proof}
\usepackage{supertabular}
\usepackage{unicode-math}
\usepackage{xltxtra}
\usepackage{xunicode}

@@

\newcommand\MkValence[3]{\left\{#1\right\}\left[#2\right]\,.\, #3}
\newcommand\MkBTm[3]{\left\{#1\right\}\left[#2\right]\,.\, #3}
\newcommand\MkArity[2]{\left(#1\right) #2}
\newcommand\IsArity[1]{\IMode{#1}\ \textit{arity}}
\newcommand\IsValence[1]{\IMode{#1}\ \textit{valence}}
\newcommand\IsSort[1]{\IMode{#1}\ \textit{sort}}
\newcommand\Leq[2]{\IMode{#1}\leq\IMode{#2}}

@@

\newcommand\ParamsOf{\pi_\mathfrak{p}}
\newcommand\ArityOf{\pi_\mathfrak{a}}
\newcommand\Sets{\mathbf{Set}}

@@

\newcommand\Grothendieck[1]{\oint #1}

@@

\newcommand\MV[1]{\textsc{#1}}
\newcommand\MApp[3]{#1\left\{#2\right\}\left(#3\right)}
\newcommand\App[2]{#1\left(#2\right)}
\newcommand\Dom[1]{\left\vert #1\right\vert}
\newcommand\Fresh[2]{\IMode{#1}\notin\IMode{\Dom{#2}}}

\newcommand\SortExp{\mathsf{exp}}

@@

Let $\Symbols$ be an infinite set of symbols; let $\Finsets$ be the free
cocartesian category over $\Symbols\times\Sorts$. Then, fix a covariant
presheaf of operators $\Operators:\Finsets\times\Arities\to\Sets$ such that the
arrows in $\Finsets$ lift to renamings of operators' parameters; via the
Grothendieck construction, we can also consider the set $\Grothendieck\Operators$ of
operators $(\Upsilon,\varrho,\vartheta)$ for
$\Member{\vartheta}{\Operators(\Upsilon,\varrho)}$.

@@

\begin{gather*}
  \infer{
    \IsMetaCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsMetaCtx{\Omega,\MV{m}:v}
  }{
    \IsMetaCtx{\Omega} &
    \IsValence{v} &
    \Fresh{\MV{m}}{\Omega}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsVarCtx{\Gamma,x:s}
  }{
    \IsVarCtx{\Gamma} &
    \IsSort{s} &
    \Fresh{x}{\Gamma}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:s}
  }{
    \IsVarCtx{\Upsilon} &
    \IsSort{s} &
    \Fresh{u}{\Upsilon}
  }
\end{gather*}

@@

\begin{gather*}
  \infer[\textit{var}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{s}
  }{
    \Lookup{\Gamma}{x}{s}
  }\\[6pt]
  %
  \infer[\textit{mvar}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{s}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{p_0,\dots,p_m}{q_0,\dots,q_n}{s}}\\
      \HypJ{\Lookup{\Upsilon}{u_i}{p_i}}{\Leq{i}{m}}\\
      \HypJ{
        \IsAbt{\Omega}{\Upsilon}{\Gamma}{M_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  %
  \infer[\textit{app}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{s}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        v_1,\dots,v_n
      }\\
      \HypJ{
        \IsBTm{\Omega}{\Upsilon}{\Gamma}{E_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  \infer[\textit{abs}]{
    \IsBTm{\Omega}{\Upsilon}{\Gamma}{
      (\MkBTm{\vec{u}}{\vec{x}}{E})
    }{
      (\MkValence{\vec{p}}{\vec{q}}{s})
    }
  }{
    \IsAbt{\Omega}{\Upsilon, \vec{u}:\vec{p}}{\Gamma,\vec{x}:{\vec{q}}}{
      E
    }{
      s
    }
  }
\end{gather*}

>>>>

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath, amsthm, stmaryrd, scalerel, amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{proof}
\usepackage{supertabular}
\usepackage{unicode-math}
\usepackage{xltxtra}
\usepackage{xunicode}

@@

\newcommand\MkValence[3]{\left\{#1\right\}\left[#2\right].\, #3}
\newcommand\MkBTm[3]{\left\{#1\right\}\left[#2\right].\, #3}
\newcommand\MkArity[2]{\left(#1\right) #2}
\newcommand\IsArity[1]{\IMode{#1}\ \textit{arity}}
\newcommand\IsValence[1]{\IMode{#1}\ \textit{valence}}
\newcommand\IsSort[1]{\IMode{#1}\ \textit{sort}}
\newcommand\Leq[2]{\IMode{#1}\leq\IMode{#2}}

@@

\newcommand\ParamsOf{\pi_\mathfrak{p}}
\newcommand\ArityOf{\pi_\mathfrak{a}}
\newcommand\Sets{\mathbf{Set}}
\newcommand\SCtx{\mathbf{SCtx}}

@@

\newcommand\Grothendieck[1]{\oint #1}
\newcommand\Comma[2]{#1 \downarrow #2}

@@

\newcommand\MV[1]{\textsc{#1}}
\newcommand\MApp[3]{#1\left\{#2\right\}\left(#3\right)}
\newcommand\App[2]{#1\left(#2\right)}
\newcommand\Dom[1]{\left\vert #1\right\vert}
\newcommand\FV[1]{\mathbf{FV}\left(#1\right)}
\newcommand\FS[1]{\mathbf{FS}\left(#1\right)}
\newcommand\NotIn[2]{\IMode{#1}\notin\IMode{#2}}
\newcommand\SortExp{\mathsf{exp}}

\newcommand\Hom[3]{#1\left[#2,#3\right]}

\newcommand\Subst[3]{
  \left[#1\, /\, #2\right]#3%
}

\newcommand\IsRenaming[3]{
  \IMode{#2}\upharpoonright\IMode{#1}%
  \leadsto\OMode{#3}%
}

\newcommand\IsSubst[4]{%
  \Subst{\IMode{#1}}{\IMode{#2}}{\IMode{#3}}%
  \leadsto\OMode{#4}%
}

\newcommand\IsEq[2]{\IMode{#1} = \IMode{#2}}
\newcommand\IsApart[2]{\IMode{#1} \mathrel{\#} \IMode{#2}}

\newcommand\Define[2]{\IMode{#1}\triangleq\OMode{#2}}

@@

Let $\Symbols$ be an infinite set of symbols. Let $\Finsets$ be the category of
finite subsets of $\Symbols$ and their injective maps; then the comma construction
$\Define{\SCtx}{\Comma{\Finsets}{\Sorts}}$, with $\Sorts$ regarded as a discrete category, is
the category of contexts of symbols, whose objects are finite sets of symbols
$U$ and sort-assignments $\mathfrak{s}:U\to\Sorts$, and whose morphisms are
sort-preserving renamings; we will write $\Upsilon$ for a symbol context
$(U,\mathfrak{s})$.

Then, fix a covariant presheaf of operators
$\Operators:\SCtx\times\Arities\to\Sets$ such that the arrows in $\SCtx$ lift
to renamings of operators' parameters; via the Grothendieck construction, we
can also consider the set $\Grothendieck\Operators$ of operators
$(\Upsilon,\varrho,\vartheta)$ for
$\Member{\vartheta}{\Operators(\Upsilon,\varrho)}$.

@@

\begin{gather*}
  \infer{
    \IsMetaCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsMetaCtx{\Omega,\MV{m}:v}
  }{
    \IsMetaCtx{\Omega} &
    \IsValence{v} &
    \NotIn{\MV{m}}{\Dom\Omega}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsVarCtx{\Gamma,x:s}
  }{
    \IsVarCtx{\Gamma} &
    \IsSort{s} &
    \NotIn{x}{\Dom\Gamma}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:s}
  }{
    \IsVarCtx{\Upsilon} &
    \IsSort{s} &
    \NotIn{u}{\Dom\Upsilon}
  }
\end{gather*}

@@

\begin{gather*}
  \infer[\textit{var}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{s}
  }{
    \Lookup{\Gamma}{x}{s}
  }\\[6pt]
  %
  \infer[\textit{mvar}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{s}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{p_0,\dots,p_m}{q_0,\dots,q_n}{s}}\\
      \HypJ{\Lookup{\Upsilon}{u_i}{p_i}}{\Leq{i}{m}}\\
      \HypJ{
        \IsAbt{\Omega}{\Upsilon}{\Gamma}{M_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  %
  \infer[\textit{app}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{s}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        v_1,\dots,v_n
      }\\
      \HypJ{
        \IsBTm{\Omega}{\Upsilon}{\Gamma}{E_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  \infer[\textit{abs}]{
    \IsBTm{\Omega}{\Upsilon}{\Gamma}{
      \left(\MkBTm{\vec{u}}{\vec{x}}{M}\right)
    }{
      \left(\MkValence{\vec{p}}{\vec{q}}{s}\right)
    }
  }{
    \IsAbt{\Omega}{\Upsilon, \vec{u}:\vec{p}}{\Gamma,\vec{x}:{\vec{q}}}{
      M
    }{
      s
    }
  }
\end{gather*}

Abstract binding trees are identified up to $\alpha$-equivalence.

%%% \subsection{Renaming of Symbols}
%%%
%%% The renaming of symbols in abstract binding trees is defined inductively by the
%%% judgment $\IsRenaming{\psi}{M}{M'}$, where
%%% $\Member{\psi}{\Hom{\SCtx}{\FS{M}}{-}}$
%%%
%%% \begin{gather*}
%%% \end{gather*}

\subsection{Substitution of Variables}

Variable substitution in abstract binding trees is defined inductively by a
pair of judgments, $\IsSubst{N}{x}{M}{M'}$ and $\IsSubst{N}{x}{E}{E'}$:

\begin{gather*}
  \infer{
    \IsSubst{N}{x}{y}{N}
  }{
    \IsEq{x}{y}
  }\qquad
  \infer{
    \IsSubst{N}{x}{y}{y}
  }{
    \IsApart{x}{y}
  }\\[6pt]
  \infer{
    \IsSubst{N}{x}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{m}}{\vec{u}}{M'_0,\dots,M'_n}
    }
  }{
    \HypJ{
      \IsSubst{N}{x}{M_i}{M'_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  \infer{
    \IsSubst{N}{x}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \App{\vartheta}{E'_0,\dots,E'_n}
    }
  }{
    \HypJ{
      \IsSubst{N}{x}{E_i}{E'_i}
    }{
      \Leq{i}{n}
    }
  }
\end{gather*}
\begin{gather*}
  \infer{
    \IsSubst{N}{x}{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{y}}{M'}
    }
  }{
    \NotIn{x}{\vec{y}} &
    \IsApart{\vec{y}}{\FV{N}} &
    \IsSubst{N}{x}{M}{M'}
  }\qquad
  \infer{
    \IsSubst{N}{x}{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }{
      \MkBTm{\vec{u}}{\vec{y}}{M}
    }
  }{
    \Member{x}{\vec{y}} &
    \IsApart{\vec{y}}{\FV{N}}
  }
\end{gather*}

Because terms are identified up to $\alpha$-equivalence, the variable
substitution judgment is functional in its inputs, and so we are justified in
writing $\Subst{N}{x}{M}$ for $M'$ when $\IsSubst{N}{x}{M}{M'}$. We write
$\Subst{\vec{N}}{\vec{x}}{M}$ for the simultaneous substitution of $\vec{N}$
for $\vec{x}$ in $M$.

\subsection{Substitution of Metavariables}

Metavariable substitution is defined inductively by the judgment
$\IsSubst{E}{\MV{m}}{M}{M'}$:

\begin{gather*}
  \infer{
    \IsSubst{E}{\MV{m}}{x}{x}
  }{
  }\\[6pt]
  \infer{
    \IsSubst{E}{\MV{m}}{
      \MApp{\MV{n}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{n}}{\vec{u}}{M'_0,\dots,M'_n}
    }
  }{
    \IsApart{\MV{m}}{\MV{n}} &
    \HypJ{
      \IsSubst{E}{\MV{n}}{M_i}{M'_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  %%% \infer{
  %%%   \IsSubst{
  %%%     \MkBTm{\vec{u}}{x_0,\dots,x_n}{N}
  %%%   }{\MV{m}}{
  %%%     \MApp{\MV{n}}{\vec{v}}{M_0,\dots,M_n}
  %%%   }{
  %%%     %TODO!!!
  %%%     %\Rename{\vec{v}}{\vec{u}}{
  %%%     \Subst{\vec{M}}{\vec{x}}{N}
  %%%     % }
  %%%   }
  %%% }{
  %%%   \IsEq{\MV{m}}{\MV{n}}
  %%% }\\[6pt]
  %%% \infer{
  %%%   \IsSubst{E}{\MV{m}}{
  %%%     \App{\vartheta}{F_0,\dots,F_n}
  %%%   }{
  %%%   }
  %%% }{
  %%% }
\end{gather*}
